---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

# Assignment
This assignment makes use of data from a personal activity monitoring
device. This device collects data at 5 minute intervals through out the
day. The data consists of two months of data from an anonymous
individual collected during the months of October and November, 2012
and include the number of steps taken in 5 minute intervals each day.

## Data

The data for this assignment can be downloaded from the course web
site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]
* The data set is in the file 'activity.csv'.

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format

* **interval**: Identifier for the 5-minute interval in which
    measurement was taken


The dataset is stored in a comma-separated-value (CSV) file and there
are a total of 17,568 observations in this
dataset.

## ESB TestDR
## Loading and preprocessing the data
```{r SCRIPT_0}
library(tibble)
library(dplyr)
acttvtee<-read.csv("./activity.csv")
tbb0 <- as_tibble(acttvtee)
str(tbb0)
tbb1 <- filter(tbb0, !is.na(steps))
```


## What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in
the dataset.(We do this above in the *Loading and preprocessing the data* step. )

### **1. Make a histogram of the total number of steps taken each day**


```{r SCRIPT_1}
tbb2<-group_by(tbb1, date) %>% summarise(step_total=sum(steps))
str(tbb2)
hist(tbb2$step_total, main = "Histogram of Total Steps a Day", xlab="Total Steps A Day")
abline(v=mean(tbb2$step_total), lwd =2, col="red")
abline(v=median(tbb2$step_total), lwd = 2, col="green")
legend("topright", legend=c("mean", "median"), col=c("red","green"), lty=1:1, lwd=4:4, cex=0.8)

```

### **2. Calculate and report the *mean* and *median* total number of steps taken per day**.

We note some summary data from the above figure.


```{r SCRIPT_2, echo=FALSE} 
print(summary(tbb2$step_total))
```  
## What is the average daily activity pattern?


### **1. Make a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**


```{r SCRIPT_3}
tbb3<-group_by(tbb1, interval) %>% summarise(avg_steps=mean(steps))
str(tbb3)
with(tbb3, plot(interval, avg_steps,type="l",xlim =c(0, 2700), xlab="time interval during day", ylab="average steps"))
title(main="Average Steps per Time Interval over 53 Days")
abline(h=mean(tbb3$avg_steps), lwd=4, col="red")
abline(h=median(tbb3$avg_steps), lwd=4, col="blue")
legend("topright", legend=c("mean", "median"), col=c("red","blue"), lty=1:1, lwd=4:4, cex=0.8)

```


***We note some summary data from the above figure.***


```{r SCRIPT_4,echo=FALSE} 
print(summary(tbb3$avg_steps))
```  
### **2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**

 The five minute interval, on average across all days in the data set that 
 contains the the maximum number of steps, is 835, as noted below:
 
```{r SCRIPT_5}
print(filter(tbb3, avg_steps==max(tbb3$avg_steps)))
```

## Imputing missing values

### **1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with `NA`s**

***Total number of missing values in the activity.csv data set is 2304 as noted below.***

```{r SCRIPT_6}
rows<-length(acttvtee$steps)
rows_of_NA_steps<-sum(is.na(acttvtee$steps))
rows_not_NA<-sum(!is.na(acttvtee$steps))
answ<-ifelse(rows== (rows_of_NA_steps + rows_not_NA),rows_of_NA_steps,NA)
answ
```

***We also note is that the field 'steps' is the only one with missing (NA) values. Note the
work which follows. We see 'date' and 'interval' have no missing values***

```{r SCRIPT_7}
rows<-length(acttvtee$date)
rows
rows_of_NA_steps<-sum(is.na(acttvtee$date))
rows_not_NA<-sum(!is.na(acttvtee$date))
answ<-ifelse(rows== (rows_of_NA_steps + rows_not_NA),rows_of_NA_steps,NA)
answ
rows<-length(acttvtee$interval)
rows
rows_of_NA_steps<-sum(is.na(acttvtee$interval))
rows_not_NA<-sum(!is.na(acttvtee$date))
answ<-ifelse(rows== (rows_of_NA_steps + rows_not_NA),rows_of_NA_steps,NA)
answ
```
### **2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.**

We will replace the missing values (which we have shown to be only in the field 'steps') with the average value
of the steps for the given time interval that was determined above.

### **3. Create a new dataset that is equal to the original dataset but with the missing data filled in.**

```{r SCRIPT_8}
t<-left_join(tbb0,tbb3)
for(i in 1:length(t$steps)){
  if(is.na(t$steps[i])) {
    t$s[i]<-t$avg_steps[i]
  }
  else {
    t$s[i]<-t$steps[i]
  }
}
t<-mutate(t, steps = s)
tbb1_1<-select(t, -avg_steps, -s)
str(tbb1_1)
```

### **4. Make a histogram of the total number of steps taken each day and Calculate and report the *mean* and *median* total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?**

***Now we will see how this impacts results found when we looked at the mean of total steps a day.***

```{r SCRIPT_9}
tbb2_1<-group_by(tbb1_1, date) %>% summarise(step_total=sum(steps))
str(tbb2_1)
hist(tbb2_1$step_total, main = "Histogram of Total Steps a Day", xlab="Total Steps A Day")
abline(v=mean(tbb2_1$step_total), lwd =2, col="red")
abline(v=median(tbb2_1$step_total), lwd = 2, col="green")
legend("topright", legend=c("mean", "median"), col=c("red", "green"), lty=1:1, lwd=4:4, cex=0.8)
```

***We note some summary data from the above figure.***


```{r SCRIPT_10, echo=FALSE} 
print(summary(tbb2_1$step_total))
```  
***And compare it with results from data where missing values were not replaced.***

```{r  SCRIPT_11, echo=FALSE} 
print(summary(tbb2$step_total))
```  


## Are there differences in activity patterns between weekdays and weekends?

Use the dataset with the filled-in missing values for this part.

### **1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.**

```{r SCRIPT_12}
tbb1_2<-tbb1_1
tbb1_2$d<-as.POSIXct(strptime(tbb1_2$date, format="%Y-%m-%d"))
tbb1_2$w<-ifelse(weekdays(tbb1_2$d)=='Saturday' | weekdays(tbb1_2$d)=='Sunday', 'Weekend', 'Weekday')
str(tbb1_2)
```
### **2. Make a panel plot containing a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).**

```{r SCRIPT_13}
library(ggplot2)
tbb3_f<-group_by(tbb1_2, w,interval) %>% summarise(avg_steps=mean(steps))
str(tbb3_f)
pl<-ggplot(tbb3_f,aes(interval, avg_steps))
pl<-pl+geom_line()
pl<-pl+facet_grid(tbb3_f$w~.)
pl<-pl+labs(title="Average Steps per Time Interval in a Day", x= "Time Interval", y="Average Steps")
print(pl)
tbb3_fwend<-filter(tbb3_f, w=="Weekend")
tbb3_fwday<-filter(tbb3_f, w=="Weekday")
```

***Let us look at the summary data for the weekend average steps per time interval.***
```{r SCRIPT_14, echo=FALSE}
summary(tbb3_fwend$avg_steps)
```
***Now we will do the same for the weekday average.***
```{r SCRIPT_15, echo=FALSE}
summary(tbb3_fwday$avg_steps)
```